\documentclass[12pt,a4paper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{natbib, hyperref, setspace,}

\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobat’s bookmarks
    pdftoolbar=true,        % show Acrobat’s toolbar?
    pdfmenubar=true,        % show Acrobat’s menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={Syllabus: Does Public Opinion Matter?},    % title
    pdfauthor={Thomas J. Leeper},     % author
    pdfsubject={Political Science},   % subject of the document
    pdfkeywords={politics} {public opinion} {political psychology}, % list of keywords
    pdfnewwindow=true,      % links in new window
    pdfborder={0 0 0}
}

\begin{document}

\section*{R problems}

\onehalfspacing

\subsection*{Math}

\begin{enumerate}
\item What is \texttt{1+NA}?
\item What is \texttt{2/0}?
\item What is \texttt{log(0)}?
\vspace{1em}
\item What is \texttt{(2+5)/10}?
\item What is \texttt{(2+5/10)}?
\end{enumerate}


\subsection*{Parsing errors}
\begin{enumerate}
\item What is \texttt{2/*1}?
\item What is \texttt{2/+1}?
\vspace{1em}
\item What is \texttt{((2+2)/(3+4)}?
\end{enumerate}


\subsection*{Variables}
\begin{enumerate}
\item What is {a} in the below:
\texttt{a <- 2
b <- 3
a * b
a <- a + b
}
\item How many items are in \texttt{c(1,NULL,2,NULL,4)}?
\vspace{1em}
\item What is \texttt{c} in the following:
\texttt{c <- 1
\lq c\lq <- 2
}
\item What happens when I type the following: \texttt{2 <- 3}?
\end{enumerate}


\subsection*{Vectors}
\begin{enumerate}
\item What is the value of \texttt{seq\_along(c(3,1,2))}?
\item Is \texttt{rep(c(1, 2), each = 4)} the same as \texttt{rep(c(1, 2), times = 4)}?
\vspace{1em}
\item What are the values of \texttt{d} and \texttt{e} in the following:
\texttt{
a <- 1:5
b <- c(1:1,3,3,2)
d <- a[b]
e <- b[a]
}
\vspace{1em}
\item What is the value of \texttt{c(1:3,4:6)[0]}?
\item What is the value of \texttt{5[1:3]}?
\vspace{1em}
\item What is the value of \texttt{1:6[rep(c(TRUE,FALSE),3)]}?
\item What is the value of \texttt{c(1,3,5,2,4,6) > 2}
\item What are the values of \texttt{b} and \texttt{c} in the following:\\
\texttt{
a <- 1:10
b <- a > 3 \& a < 8
c <- a > 3 \& a <= 2
}
\end{enumerate}


\section*{Matrices}
\begin{enumerate}
\item What is the value of \texttt{a[,1]} in the following:\\
\texttt{a <- matrix(1:6, nrow=2)}
\item What is the value of {a[2,]} (from previous)?
\end{enumerate}


\section*{Dataframe Structure and indexing}
\begin{enumerate}
\item What are the dimensions of the \texttt{Orange} dataframe?
\item What are the names of the variables in the \texttt{iris} dataframe?
\item What are the rownames of the \texttt{iris} dataframe?
\item What is the first row of data from the \texttt{PlantGrowth} dataframe?
\end{enumerate}


\section*{Data summaries}
\begin{enumerate}
\item What is the mean \texttt{weight} in the \texttt{PlantGrowth} dataframe?
\item How many unique values are there for the \texttt{group} variable in the \texttt{PlantGrowth} dataframe?
\item What is the range of the \texttt{weight} variable in the \texttt{PlantGrowth} dataframe?
\item What is the correlation between \texttt{height} and \texttt{weight} in the \texttt{women} dataframe?
\item How many \texttt{spontaneous} abortions are reported at each level of \texttt{education} in the \texttt{infert} dataframe?
\item Which state had the highest \texttt{Murder} rate in the \texttt{USArrests} dataset?
\end{enumerate}

\section*{Plotting}
\begin{enumerate}
\item How would we draw a scatterplot of \texttt{height} and \texttt{weight} for the \texttt{women} dataframe?
\item How would we make it red?
\item How would we make a boxplot of \texttt{circumference} by \texttt{Tree} in the \texttt{Orange} dataframe?
\item How would we plot the \texttt{anscombe} sets?
\end{enumerate}


\section*{Scale construction}
\begin{enumerate}
\item Combining \texttt{spontaneous} and \texttt{induced}, what is the mean number of abortions for individuals in the \texttt{infert} dataframe?
\end{enumerate}


\section*{Regression}
\begin{enumerate}
\item What is the effect of \texttt{UrbanPop} on \texttt{Murder} rate in the \texttt{USArrests} dataframe?
\item What are the effects of \texttt{Catholic} and \texttt{Education} on \texttt{Fertility} in the \texttt{swiss} dataframe? Is there an interaction?
\item What are the effects of \texttt{supp} and \texttt{dose} on \texttt{len} in the \texttt{ToothGrowth} dataframe? Is there an interaction?
\end{enumerate}


\section*{Regression plotting}
\begin{enumerate}
\item How would you draw a plot of the effect of \texttt{speed} on stopping \texttt{distance} in the \texttt{cars} dataframe?
\item How would we plot the effects (on \texttt{len}) of \texttt{dose} for each type of \texttt{supp} in the \texttt{ToothGrowth} dataframe?
\end{enumerate}

\end{document}